### 2018-06-02 doc/manuscript.Rmd
add, rearrange discussion items about dnl and alt explain

There are some possible explanations to these findings.

{{
There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
(DNL). Given that most fatty acids can come from either source, it is difficult
to

}}

Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
DNL may influence metabolic dysfunction is not well described. One possible explanation
may be the relationship between greater DNL and the types of fatty acids produced.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TGFA in circulation
[@Chehade2013a; @Verges2015a]. We found that
higher proportions of these four DNL TGFA also highly correlated with a higher
concentration of clinical TG, reinforcing this pathway between carbohydrate
intake, DNL, and circulating TG. The higher concentration of circulating 14 and
16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
instance from palmitic acid (16:0), which is well-known to have harmful effects
on tissues [@Riserus2008a; @Iggman2010a].

@@

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In a previous published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].

<!--
{{ comment on comparing models with and without NEFA }} While NEFA may be a
mediating factor, we found that adjustment for it did not attenuate this
association since TGFA composition is also dictated by other sources for DNL
(e.g. recent meal, DNL from glucose stores, fatty acid recycling, etc.).

- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}

>>>>

There are a few possible explanations to these findings. The composition of the
TGFA fraction is determined by the two sources of dietary fat and de novo
lipogenesis (DNL). And while adipose tissue also supplies circulating fatty
acids, the stored fatty acids ultimately originate from these two sources.
Specific dietary fat as well as carbohydrates can influence DNL activity
{{cite}}. Since the majority of fatty acids can be obtained from either source,
the specific source for fatty acids is extremely difficult to ascertain outside
of highly controlled experimental settings. 
In our findings, the four fatty acids are also highly correlated amongst each other, in addition to
clustering together on their negative association with insulin sensitivity. These
findings suggest that these fatty acids may derived mainly from DNL, given the
evidence that they share a tight DNL pathway. While it may be possible that
these fatty acids also share a common dietary source (e.g. red meat) {{cite}},
dietary sources of these four fatty acids are also high in fatty acids such as
18:0 and 18:1 n-9, which we did not see cluster together.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. {{talk about Yki-.. study}}. These fatty acids are similar
to the fatty acids we found that clustered together using the PLS analysis.

We found that higher proportions of these four DNL TGFA also highly correlated
with a higher concentration of clinical TG, highlighting a potential relationship
between these four fatty acids, circulating TG, and the

A higher activity in DNL may be a means to control blood glucose with higher
dietary carbohydrate intake [@Hodson2008a; @Kawano2013a; @Harding2015a;
@Hudgins2000a; @Parks1999a; @Chehade2013a; @Verges2015a], which could increase
hepatic fat stores and increase the amount of TG in circulation. Several studies
have in fact shown a link between higher estimated DNL and an increased risk for
metabolic dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a];
however, how DNL may influence metabolic dysfunction is not well described. One
possible explanation may be the relationship between greater DNL and the types
of fatty acids produced.

Regardless of the exact source of these fatty acids, our results in addition to
the available scientific evidence emphasizes the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have unique
physiological functions {{cite}}.

The higher concentration of
circulating 14 and 16 carbon fatty acids may then expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a].

@@

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].

<!--
- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}
